Bootstrapping Multi-atlas Hippocampal Segmentation with MAGeT

### 2013-03-28 paper.Rnw
Updates to the results section...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, 
    fig.align = 'center', tidy = FALSE, comment = NA, cache = TRUE)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw())
# bland_altman_plot <- function(dat){
#   dat$diff <- dat$pre - dat$post;
#   
#   return ggplot(dat, aes(pre, diff)) + 
#     geom_hline(yintercept = mean(dat$diff) + c(-2, 0, 2) * sd(dat$diff), 
#                linetype=2, color='brown') + 
#     geom_point() 
# }
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

all_data = merge(ma, mb, by=c("timestamp", "atlases", "batch", "label", "reg_method", "subject"))
all_data$k_diff = all_data$k.y - all_data$k.x 
method_labels <- list("Majority Vote" = "majvote", 
                      "NMI Vote" = "nmi", 
                      "Cross-correlation Vote" = "xcorr")
levels(all_data$method.y) <- method_labels
all_data       = merge(all_data, diagnoses)
remove(t_ma, a_ma, t_mb, a_mb, ma, mb)
@
%                                                                              %
%                                                                              %
 % Bland-Altman plot for MA and MAGeT

\begin{figure}
<<ADN1-Xval-k-mean, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
ggplot(all_data, aes(x=templates.y,y=k.y, colour=as.factor(atlases))) + 
  facet_grid(reg_method~method.y) + 
  #stat_summary(fun.y=mean,geom='line',
  #             aes(y=k.y)) + 
  stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" )
@
  \caption{MAGeT brain mean DSC by atlas and template library size}
  \label{fig:ADNI1_xval-k_mean}
\end{figure}

@@

@@

ggplot(stddev.t.test, 
       aes(x=templates.y, y=p.value, colour=as.factor(atlases))) + 
  geom_line() + 
  geom_hline(aes(yintercept=0.05), linetype='dashed') + 
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_x_continuous(breaks=1:20) + scale_y_continuous(breaks=seq(0,1,by=0.05)) + 
  scale_y_log10() + 
  xlab( "Number of Templates" ) + 
  ylab( "p-value" ) 
@
\end{figure}

@@

\begin{figure}
<<ADN1-Xval-ANTS-20tmpls-Bland-Atlman, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
a_20_tmpls = subset(all_data, 
                      reg_method  == "ANTS" &
                      method.y    == "Majority Vote" &
                      templates.y == 20 &
                      atlases     == 9 &
                      volume.x < 10000)
avg = ddply(a_20_tmpls, c("subject", "label"), function (df) {
  data.frame(
    pre = mean(df$volume.x),  #MA
    post = mean(df$volume.y)  #MB
  )
})
# TODO: against SNT

#df = data.frame(pre=a_20_tmpls$volume.x, post=a_20_tmpls$volume.y)
bland_altman_plot(avg) + 
  xlab('Mean of MA and MAGeT HC Left/Right Volumes') + 
  ylab('MA - MAGeT Volume') + 
  opts(title = 'Bland-Altman Plot')
@
  \caption{Change in mean Kappa between multi-atlas and MAGeT brain using ANTS, 20 templates}
  \label{}
\end{figure}

\begin{figure}
<<ADN1-Xval-k-by-disease, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
a_20_tmpls = subset(all_data, 
                      reg_method  == "ANTS" &
                      method.y    == "Majority Vote" &
                      templates.y == 20 &
                      atlases %in% c(1,9))
template_library = read.csv('data/a2a_ants/template_library_composition.csv')
template_library$subject = NULL
a_20_tmpls = merge(a_20_tmpls, template_library,
                   by.x=c('timestamp', 'templates.y'), 
                   by.y=c('timestamp', 'templates'))
ggplot(a_20_tmpls, aes(y=k.y, x=templates_CN,colour=DX)) +  # why does k.x look the same?
  facet_grid( . ~ atlases) + 
  geom_smooth(method='lm',fullrange=F) 
 # add 
@
  \caption{Mean kappa vs. proportion of disease category in template library by disease category}
  \label{}
\end{figure}

- find significant improvement over multi-atlas performed with the same
  parameters. Also, find smoothed performance is monotonically increasing but
  asymptotic in size of both template and atlas library, with peak performance
  reached after 15 templates.

- True for all parameter settings! (so works generally..)
\todo{Can we statistically capture what "peak" performance is?  Something like,
the point at which gains become statistically insignificant?}

Ideas:
\begin{itemize}
\item more atlases -> better performance
\item  larger template library -> better performance, but tails off around 10-15 templates
\item  no significant difference between majority or weighted vote methods (haven???t tested this statistically though). 
\item  consistently performs better than average naive performance by XXX
\item  using ANTS, with a large enough template library (>12) MAGeT brain performs
  better than the average multi-atlas approach with the same number of atlases.
  using ANIMAL, 5 or more atlases needed before boost seen. 
\item  more atlases -> smaller template library required to improve on average multi-atlas performance
\item  discuss variance?  best/worst case? --how often do we expect random template library selection to work decently
\end{itemize}

\todo{show cost (in registrations) / benefit tradec off graph:  show number of
registrations per Kappa?  or hours of manual labour per Kappa?)}
% discuss run-time for each of MAGet and multi-atlas

%%%%%
%%%%%                  Results: ADNI-1 Screening Validation               %%%%%
%%%%%
\subsection{ADNI-1 Screen Validation}

Ideas: 
\begin{itemize}
\item  A2A shows that if atlas population strongly(?) represents subject set
variability, then free choice from atlas population will produce improvements
(we know this b/c of extensive validation trials). 
\item  what about in the case where atlas population doesn???t strongly
represent subject set variability (e.g. a priori atlas set)?  then, we can use
atlas selection to refine atlas set? 
\end{itemize}

\todo{Kappa against our manual rater is low... explain that}

@@

% paired t-test against SNT to show similarity (should be low) 
% Kruskal-Willis against  (should be high?)
% post-hoc power analysis to say we can detect this change
\begin{figure}[h]
<<ADNI-scr-histogram,dependson='ADNI-baseline-volumes-boxplot',cache=T>>=
melted   = melt(totals.complete, measure.vars=c( "SNT","FSL","FS","MAPER","MAGeT"))
binwidth = 250

@@

label_x_pos = 10000
ggplot(data=melted, aes(x=value)) + 
  geom_histogram(binwidth=binwidth, alpha=0.8) + 
  geom_line(aes(y=freq, colour='red'), data=normaldf) +
  geom_text(aes(y=29,x=label_x_pos,hjust=1,size=1, 
                label=paste("list(W ==",signif(shapiro.w, digits=3), ", p == ", 
                            signif(shapiro.p,digits=3), ")")), 
            parse=TRUE, data=testsdf) +
  geom_text(aes(y=25,x=label_x_pos,hjust=1,size=1, 
                label=paste("list(D ==",signif(ks.D, digits=3), ", p == ", 
                            signif(ks.p,digits=3), ")")), 
            parse=TRUE, data=snt.tests) + 
  geom_text(aes(y=21,x=label_x_pos,hjust=1,size=1, 
                label=paste("sigma ==",signif(sd, digits=3))), 
            parse=TRUE, data=testsdf) + 
  geom_text(aes(y=17,x=label_x_pos,hjust=1,size=1, 
                label=paste("mu ==",signif(mean, digits=3))), 
            parse=TRUE,data=testsdf) + 
  xlim(0, 10000) + ylim(0,30) + 
  facet_grid(variable ~ DX) + 
  scale_colour_discrete(legend=FALSE) + 
  scale_size_continuous(legend=FALSE) + 
  labs(title="Total hippocampus volume by disease and measurement method", 
       x=expression(paste("Total hippocampal volume (", mm^2, ")")),
       y="Frequency") + 
  opts(axis.title.x = theme_text(vjust=-0.1))
@
  \caption{{\bf ADNI Baseline cohort.} Comparison of total hippocampal volumes as measured by SNT, FSL, FreeSurfer (FS), MAPER, and MAGeT brain (MAGeT). A fitted normal curve is shown in red. $W$ is the Shapiro-Wilkes test statistic measuring normality of the data (signficance indicates non-normality). $D$ is the Kolmogorov-Smirnov test statistic measuring the goodness-of-fit between the distribution of measured volumes and SNT volumes (significance indicates a difference).}
  \label{ADNI-scr-histogram}
\end{figure}

@@

% show as a plot
<<ADNI-scr-summary-table, echo=F, results="asis",cache=FALSE>>=
# a rather dirty way of computing the pairwise correlations and then forming
# a printable table out of them.  Sorry.
melted = melt(totals.complete, measure.vars=c("SNT","FSL", "FS", "MAPER"))
cor_by_dx    = by(melted[c("MAGeT", "value")], c(melted['variable'], melted['DX']), 
                  function(x) cor(x$MAGeT,x$value, use="pairwise.complete.obs"))
cor_overall  = by(melted[c("MAGeT", "value")], melted['variable'], 
                  function(x) cor(x$MAGeT,x$value, use="pairwise.complete.obs"))
cor_by_dx = t(rbind(cor_by_dx))
cor_overall = rbind(cor_overall)

@@

latex(df, title="",
      n.cgroup=c(1,ncol(df)-1),cgroup=c("","Volume Correlation"),
      )#file="", size="scriptsize")
@

@@

<<WAval-error,cach=T>>=
# compare % error between MAGeT and resampling
mb0.6mm_diff_over_gold_diff = cbind(subjects, abs(gold_diff) - abs(mb_0.6mm_diff), res="diffdff")
names(mb0.6mm_diff_over_gold_diff) <- c("subject", "CA1", "Subiculum", "CA4", "CA2/CA3", "Dentate", "res")
melted = melt(mb0.6mm_diff_over_gold_diff, id.vars=c("subject", "res"))
names(melted) <- c("subject","res", "region","voldiff")
ggplot(melted, aes(y=voldiff,x=region)) +
  geom_boxplot() +
  labs(title="Absolute difference in % difference between MAGeT and resampling",
       y="Absolute difference in % difference in volume", 
       x="Region")
@
<<WAval-kappa,cache=T>>=
data = read.csv('data/WAval-kappa.csv')
ggplot(data, aes(x=label, y=kappa)) + geom_boxplot(y=kappa)
@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
#opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, 
#    fig.align = 'center', tidy = FALSE, comment = NA, cache = TRUE)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw())
# bland_altman_plot <- function(dat){
#   dat$diff <- dat$pre - dat$post;
#   
#   return ggplot(dat, aes(pre, diff)) + 
#     geom_hline(yintercept = mean(dat$diff) + c(-2, 0, 2) * sd(dat$diff), 
#                linetype=2, color='brown') + 
#     geom_point() 
# }
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

all_data = merge(ma, mb, by=c("timestamp", "atlases", "batch", "label", "reg_method", "subject"))
all_data$k_diff = all_data$k.y - all_data$k.x 
method_labels <- list("Majority Vote" = "majvote", 
                      "NMI Vote" = "nmi", 
                      "Cross-correlation Vote" = "xcorr")
levels(all_data$method.y) <- method_labels
all_data       = merge(all_data, diagnoses)
remove(t_ma, a_ma, t_mb, a_mb, ma, mb)
@
%                                                                              %
%                                                                              %

% TODO: Bland-Altman plot for MA and MAGeT

Hippocampal MAGeT-based segmentations using both ANIMAL and ANTs registration
algorithm demonstrate good overlap with manually derived gold-standards ( Figure
\ref{fig:ADNI1-xval-k-mean}, \todo{table?}) using each label fusion method.
Qualitatively, both ANIMAL and ANTs-based segmentations demonstrate trend
overlap accuracy that increases with the size of atlas library and template
library, with at least 12 templates needed for optimal results
\todo{quantify?}).  No significant difference in segmentation accuracy is seen
when either ANIMAL or ANTs registration is used with different label fusion
techniques, at any atlas or template library sizes (\todo{lm of interaction?).
In every parameter configuration, the use of MAGeT Brain with ANTs registration
shows increased segmentation accuracy over MAGeT Brain with ANIMAL registration
(\todo{interaction?}).

It is interesting to note that with an even number of templates, MAGeT brain
shows a small decrease in performance relative to when one fewer template image
is used (\todo{show significance of this}). See section (\todo{ref}) for a
discussion of this behaviour. In the remainder of this section, only results
from odd-sized template libraries will be shown.

\begin{figure}
<<ADN1-xval-k-mean, cache=TRUE, dependson='ADNI1-Xval-prep'>>=

ggplot(subset(all_data, method == 'majvote'), 
       aes(x=templates.y,y=k.y, colour=as.factor(atlases))) + 
  facet_grid(reg_method~method.y) + 
  #stat_summary(fun.y=mean,geom='line',
  #             aes(y=k.y)) + 
  stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" )
@
  \caption{MAGeT brain mean DSC by atlas and template library size}
  \label{fig:ADNI1_xval-k_mean}
\end{figure}

With an increasing number of templates, MAGeT-based segmentating using ANTs
registration and majority vote label fusion shows improvement in overlap
accuracy over multi-atlas-based segmentation, using the same number of atlases
and voting method Figure \todo{ref}. The magnitude of improvement over
multi-atlas-based segmentation decreases with an increasing number of atlases,
with accuracy converging with 7 atlases.  Peak improvement in MAGeT Brain
accuracy (0.02 DSC) is found when one atlas is used with a template library of
20 images. \todo{stats?}

@@

In addition to an improvement in accuracy over multi-atlas-based segmentation,
MAGeT Brain also shows a decrease in the variability of segmentation accuracy
with different input atlases (Figure \todo{ref}).  Additionally, variability
decreases with the size of template library, and is significantly decreased
(p<0.05) when using 3-7 atlases and 17 or more templates.

@@

ggplot(stddev.t.test, 
       aes(x=templates.y, y=p.value, colour=as.factor(atlases))) + 
  geom_line() + 
  geom_hline(aes(yintercept=0.05), linetype='dashed') + 
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_x_continuous(breaks=1:20) + scale_y_continuous(breaks=seq(0,1,by=0.05)) + 
  xlab( "Number of Templates" ) + 
  ylab( "p-value" ) 
@
\end{figure}

@@

% \begin{figure}
% <<ADN1-Xval-ANTS-20tmpls-Bland-Atlman, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
% a_20_tmpls = subset(all_data, 
%                       reg_method  == "ANTS" &
%                       method.y    == "Majority Vote" &
%                       templates.y == 20 &
%                       atlases     == 9 &
%                       volume.x < 10000)
% avg = ddply(a_20_tmpls, c("subject", "label"), function (df) {
%   data.frame(
%     pre = mean(df$volume.x),  #MA
%     post = mean(df$volume.y)  #MB
%   )
% })
% # TODO: against SNT
% 
% #df = data.frame(pre=a_20_tmpls$volume.x, post=a_20_tmpls$volume.y)
% bland_altman_plot(avg) + 
%   xlab('Mean of MA and MAGeT HC Left/Right Volumes') + 
%   ylab('MA - MAGeT Volume') + 
%   opts(title = 'Bland-Altman Plot')
% @
%   \caption{Change in mean Kappa between multi-atlas and MAGeT brain using ANTS, 20 templates}
%   \label{}
% \end{figure}
% 
% 
% \begin{figure}
% <<ADN1-Xval-k-by-disease, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
% a_20_tmpls = subset(all_data, 
%                       reg_method  == "ANTS" &
%                       method.y    == "Majority Vote" &
%                       templates.y == 20 &
%                       atlases %in% c(1,9))
% template_library = read.csv('data/a2a_ants/template_library_composition.csv')
% template_library$subject = NULL
% a_20_tmpls = merge(a_20_tmpls, template_library,
%                    by.x=c('timestamp', 'templates.y'), 
%                    by.y=c('timestamp', 'templates'))
% ggplot(a_20_tmpls, aes(y=k.y, x=templates_CN,colour=DX)) +  # why does k.x look the same?
%   facet_grid( . ~ atlases) + 
%   geom_smooth(method='lm',fullrange=F) 
%  # add 
% @
%   \caption{Mean kappa vs. proportion of disease category in template library by disease category}
%   \label{}
% \end{figure}
% 
% 
% \todo{show cost (in registrations) / benefit tradec off graph:  show number of
% registrations per Kappa?  or hours of manual labour per Kappa?)}
% discuss run-time for each of MAGet and multi-atlas

%%%%%
%%%%%                  Results: ADNI-1 Screening Validation               %%%%%
%%%%%
\subsection{ADNI-1 Screening Validation}

Based on the results from the ADNI1 Cross-Validation experiment, MAGeT Brain was
configured with a template library of 20 subject images and majority vote label
fusion.

% \begin{itemize}
% \item  A2A shows that if atlas population strongly(?) represents subject set
%        variability, then free choice from atlas population will produce
%        improvements (we know this b/c of extensive validation trials). 
% \item  what about in the case where atlas population doesn???t strongly
%        represent subject set variability (e.g. a priori atlas set)?  then, we
%        can use atlas selection to refine atlas set? 
% \end{itemize}
% 
% \todo{Kappa against our manual rater is low... explain that}

@@

% paired t-test against SNT to show similarity (should be low) 
% Kruskal-Willis against  (should be high?)
% post-hoc power analysis to say we can detect this change
\begin{figure}[h]
<<ADNI-scr-histogram,cache=T>>=
melted   = melt(totals.complete, measure.vars=c( "SNT","FSL","FS","MAPER","MAGeT"))
binwidth = 250

@@

label_x_pos = 10000
ggplot(data=melted, aes(x=value)) + 
  geom_histogram(binwidth=binwidth, alpha=0.8) + 
  geom_line(aes(y=freq, colour='red'), data=normaldf) +
  geom_text(aes(y=29,x=label_x_pos,hjust=1,size=1, 
                label=paste("list(W ==",signif(shapiro.w, digits=3), ", p == ", 
                            signif(shapiro.p,digits=3), ")")), 
            parse=TRUE, data=testsdf) +
  geom_text(aes(y=25,x=label_x_pos,hjust=1,size=1, 
                label=paste("list(D ==",signif(ks.D, digits=3), ", p == ", 
                            signif(ks.p,digits=3), ")")), 
            parse=TRUE, data=snt.tests) + 
  geom_text(aes(y=21,x=label_x_pos,hjust=1,size=1, 
                label=paste("sigma ==",signif(sd, digits=3))), 
            parse=TRUE, data=testsdf) + 
  geom_text(aes(y=17,x=label_x_pos,hjust=1,size=1, 
                label=paste("mu ==",signif(mean, digits=3))), 
            parse=TRUE,data=testsdf) + 
  xlim(0, 10000) + ylim(0,30) + 
  facet_grid(variable ~ DX) + 
  scale_colour_discrete(guide="none") + 
  scale_size_continuous(guide="none") + 
  labs(title="Total hippocampus volume by disease and measurement method", 
       x=expression(paste("Total hippocampal volume (", mm^2, ")")),
       y="Frequency") + 
  theme(axis.title.x = element_text(vjust=-0.1))
@
  \caption{{\bf ADNI Baseline cohort.} Comparison of total hippocampal volumes
  as measured by SNT, FSL, FreeSurfer (FS), MAPER, and MAGeT brain (MAGeT). A
  fitted normal curve is shown in red. $W$ is the Shapiro-Wilkes test statistic
  measuring normality of the data (signficance indicates non-normality). $D$ is
  the Kolmogorov-Smirnov test statistic measuring the goodness-of-fit between
  the distribution of measured volumes and SNT volumes (significance indicates a
  difference).}
  \label{ADNI-scr-histogram}
\end{figure}

@@

% show as a plot
<<ADNI-scr-summary-table, echo=F, results="asis",cache=FALSE>>=
# a rather dirty way of computing the pairwise correlations and then forming
# a printable table out of them.  Sorry.
melted = melt(totals.complete, measure.vars=c("MAGeT","FSL", "FS", "MAPER"))
cor_by_dx    = by(melted[c("SNT", "value")], c(melted['variable'], melted['DX']), 
                  function(x) cor(x$SNT,x$value, use="pairwise.complete.obs"))
cor_overall  = by(melted[c("SNT", "value")], melted['variable'], 
                  function(x) cor(x$SNT,x$value, use="pairwise.complete.obs"))
cor_by_dx = t(rbind(cor_by_dx))
cor_overall = rbind(cor_overall)

@@

latex(df, title="",
      n.cgroup=c(1,ncol(df)-1),cgroup=c("","Volume Correlation"),
      file="", size="scriptsize")
@

@@

<<WAval-error,cach=T>>=
# compare % error between MAGeT and resampling
mb0.6mm_diff_over_gold_diff = cbind(subjects, abs(gold_diff) - abs(mb_0.6mm_diff), res="diffdff")
names(mb0.6mm_diff_over_gold_diff) <- c("subject", "CA1", "Subiculum", "CA4", "CA2/CA3", "Dentate", "res")
melted = melt(mb0.6mm_diff_over_gold_diff, id.vars=c("subject", "res"))
names(melted) <- c("subject","res", "region","voldiff")
ggplot(melted, aes(y=voldiff,x=region)) +
  geom_boxplot() +
  labs(title="Absolute difference in % difference between MAGeT and resampling",
       y="Absolute difference in % difference in volume", 
       x="Region")
@
<<WAval-kappa,cache=T>>=
#data = read.csv('data/WAval-kappa.csv')
ggplot(data, aes(x=label, y=kappa)) + geom_boxplot(y=kappa)
@
